Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-09 paper/geometry-twocolumn.tex
Tighten up the general framework section
Patty's changes

We motivate this paper's goal of investigating algebraically-indexed
types and their relational interpretations by presenting a novel type
system for writing programs that manipulate two-dimensional geometric
data. Geometry is rich with operations that are invariant under
transformation: affine operations are invariant under change of
origin, vector space operations are invariant under change of basis,
and the Euclidean dot product is invariant under orthogonal changes of
basis. Moreover, some operations are interestingly variant under
transformations: for example, the cross product varies with scalings
of the plane. We incorporate this extra information about our
geometric primitives into the type system by means of algebraically
indexed types.

\subsection{Origin Invariance and Representation Independence}

When writing programs that manipulate geometric data, the basic data
structure is the $n$-tuple of numbers. In the 2-dimensional case,
tuples $\vec{v} = (x,y)$ are called upon to represent both
\emph{points}---offsets from some origin---and
\emph{vectors}---offsets in their own right. Points and vectors are
very different despite their common representation.

Distinguishing between points and vectors is the key feature of
\emph{affine geometry} (see, for example, Gallier's book
\cite{gallier11geometric}, Chapter 2). Traditionally in computational
geometry libraries, the distinction between points and vectors has
either been left to the programmer to maintain by themselves, or has
been enforced through the use of separate abstract types for points
and vectors (one might say that the standard mathematical formulation
of affine spaces uses the abstract types approach). Here, we
investigate a more sophisticated approach using types indexed by
change of origin transformations, and think of the difference between
points and vectors as a change of data representation.

Putting this in concrete terms, if we have two origins, $(0,0)$ and
$(10,20)$, then the tuple $(1,1)$ with respect to the first origin and
$(11,21)$ with respect to the second origin represent the \emph{same}
point because they have the same displacement from their respective
origins. It is merely an artifact of our representation of points as
pairs of numbers that these appear to be different. When we write
programs that manipulate data that represents points, we ought to
ensure that our programs are invariant with respect to changes in the
choice of origin. On the other hand, vectors are not intended to be
regarded as invariant under change of origin. Vectors represent
offsets and the vector $(0,0)$ always represents the zero offset.

Invariance under change of representation immediately recalls
Reynolds' fable about two professors teaching the theory of complex
numbers \cite{reynolds83types}. One professor chooses rectangular
coordinates ($x + iy$), while the other chooses polar coordinates
($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after learning the
basic operations on complex numbers in the two representations, the
two classes can interact because the theory of complex numbers is
invariant with respect to the choice of representation. Reynolds
formalises the idea of invariance with respect to changes of
representation as preservation of relations. If a binary relation $R$
relates two representations of a concept---for example the rectangular
and polar representations of the same complex number are
related---then a program that manipulates complex numbers at a level
of abstraction above their specific representation should preserve the
relation $R$. For example, let $f$ be a program taking complex numbers
to complex numbers that respects invariance under change of
representation. Then if $c$ is a complex number in rectangular form,
and $c'$ is a complex number in polar form such that $R$ relates them,
which we write as $(c,c') \in R$, then it is the case that $(f(c),
f(c')) \in R$.

We now apply Reynolds' relational approach to the geometric
setting. We define a family of binary relations on $\mathbb{R}^2$
which is indexed by changes of origin. Changes of origin are
represented by vectors in $\mathbb{R}^2$, and form a group under
addition. To emphasise the intended meaning as translations, we write
the collection of all change of origin offsets as $\Transl{2}$. By
quantifying over all changes of origin we will be able to ensure that
our programs are invariant with respect to the particular choice of
origin. The members of the $\Transl{2}$-indexed family of binary
relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
\}_{\vec{t} \in \Transl{2}}$ are defined as follows:
\begin{displaymath}
  R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}
\end{displaymath}

Now consider a function $f$ that, for example, takes two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that all these tuples represent points with respect to the same
origin, but that the function itself should be invariant with respect
to the choice of origin. We use Reynolds' idea of preservation of
relations to formalise this: for any $\vec{t} \in \Transl{2}$ (i.e.~any
change of origin), the function $f$ should satisfy the following
statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}}, (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}), f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
\end{equation}
Unfolding the definition of $R_{\vec{t}}$, this statement is
equivalent to the following statement expressing invariance of $f$
with respect to changes of origin: for all $\vec{t} \in \Transl{2}$:
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} + \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
\end{equation}
So Reynolds' preservation of relations, when instantiated with the
family of relations $\{R_{\vec{t}}\}$, yields exactly the geometric
property of invariance under change of origin.

@@

Reynolds' further insight was to see how preservation of relations by
the denotational interpretations of programs can automatically be
established by following a typing discipline. In Reynolds' case, the
typing discipline was the polymorphic $\lambda$-calculus where new
types can be constructed by universal quantification over all
types. Universal type quantification is commonly written $\forall
\alpha. A$, which denotes a type that may be instantiated with any
other type $B$ by replacing the type variable $\alpha$ with $B$ in
$A$.

In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. In the statements of geometric invariance that we stated in
the previous section (e.g., \statementref{eq:f-invariant-frame}), we
did not quantify over all relations. Instead, we quantified over all
changes of origin and used the choice to select a relation from the
family $\{R_{\vec{t}}\}$. Driven by this observation, we introduce
quantification over changes of origin into the language of types.  We
use the notation $\forall t \mathord: \SynTransl{2}. A$ for
quantification over all 2-dimensional translations (i.e.~choices of
origin) $t$. We refer to $\SynTransl{2}$ as the \emph{sort} of
$t$. Note that we have used a different font to distinguish the
semantic group $\Transl{2}$ from the syntactic sort $\SynTransl{2}$.

The sort $\SynTransl{2}$ represents an abelian group, so we can combine
its elements using the usual group operations. We write operations
additively, using $e_1 + e_2$ for the group operation, $-e$ for
inverse and $0$ for the unit.  We also regard expressions built from
variables and the group operations up to the abelian group axioms. For
example, $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) + e_3$ are to be
regarded as equivalent.

Our language of types also includes the common type constructors for
function types, $A \to B$, sum types $A + B$, unit type
$\tyPrimNm{unit}$ and tuple types $A \times B$. We also assume a
primitive type $\tyPrimNm{real}$, used to represent scalars. As we
noted at the start of the previous section, the central data structure
in geometric applications is the tuple of real numbers for
representing points and vectors. However, we cannot simply express
this as the type $\tyPrimNm{real} \times \tyPrimNm{real}$ because it
will not have the correct relational interpretation. (Two elements of
type $\tyPrimNm{real}$ will be related if and only if they are equal,
and hence two elements of $\tyPrimNm{real} \times \tyPrimNm{real}$
will be related if and only if they are equal, by Reynolds' definition
of the relational interpretation of tuple types.) Therefore, we
introduce a new type indexed by expressions $e$ of sort
$\SynTransl{2}$ to represent points \emph{and} vectors:
$\tyPrim{vec}{e}$. The indexing argument $e$ represents the possible
amount of displacement by change of origin this point may have been
subjected to. We have used the name $\tyPrimNm{vec}$, even though we
have taken pains to separate geometric points and vectors, to recall
the computer science notion of vector as a sequence of values of
homogeneous type which has a known length (in this case two).

We now sketch the interpretation of the types $\tyPrim{vec}{e}$ and
$\forall t\mathord:\SynTransl{2}.A$. In our calculus, each type has two
interpretations: an index-erasure interpretation that ignores the
indexing expression $e$, and a relational interpretation as a binary
relation on the index-erasure interpretation. For now, we assume we
are given some way of turning an expression $e$ of sort $\SynTransl{2}$
into an actual element $\sem{e}\rho$ of the group $\Transl{2}$, under
some environment $\rho$ which tells how to interpret the free
variables. The (simplified) index-erasure and relation interpretations
of the types $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure interpetation of universal quantification $\forall
t\mathord:\SynTransl{2}.A$ is simply the interpretation of the
underlying type $A$. The relational interpretation intersects the
relational interpretations of the type $A$ under all extensions of the
relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A} \\
    \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{ \rsem{A}{}{\rho'} \sepbar \rho' \in \extends{\rho}{t\mathord:\SynTransl{2}} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{t\mathord:\SynTransl{2}}$
depends on which (possibly non-compositional) interpretation of
relational environments that we take. For now, it will suffice to
think of relational environments as simply mappings from free
variables of sort $\SynTransl{2}$ to the group $\Transl{2}$, and
extension as just adding an extra component to the tuple. The general
index-erasure semantics will be given in
\autoref{sec:index-erasure-semantics}. The relational interpretation
is in general more complex, due to the possibility of
non-compositional interpretations of free index variables. The
relational interpretation will be presented in
\autoref{sec:relational-semantics}.

In the previous section, we discussed functions $f : \mathbb{R}^2
\times \mathbb{R}^2 \to \mathbb{R}^2$ that have the property that they
preserve all changes of origin. We can now express this property as a
type:
\begin{displaymath}
  \mathrm{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}
\end{displaymath}
Spelling out the relational interpretation of this type using the
definitions above (and the standard relational interpretations for
tuple and function types), we obtain
\statementref{eq:f-invariant-frame}.

@@

Invariance under change of origin is the key feature of affine
geometry, and the central operation of affine geometry is the affine
combination of points: $\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$,
where $\lambda_1 + \lambda_2 = 1$. Geometrically, affine combination
can be interpreted as interpolation between the points represented by
$\vec{v_1}$ and $\vec{v_2}$. We add affine combination of points to
our calculus, with the following typing and intended denotation:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affComb} :\forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\
    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} + r\vec{v_2}
  \end{array}
\end{displaymath}
It can be verified by hand that the intended denotation
$\sem{\mathrm{affComb}}$ has the invariance property specified by its
type.
% By defining the primitive function $\mathrm{affComb}$ to take a
% single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
% are taking the affine combination of two representatives of points,
% and not just an arbitrary linear combination.

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive and three steps of De
  Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t} \\
      \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\
      \quad \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2) \\
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$ and a control point $p_1$, an
  application $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ for $s
  \in [0,1]$ gives us the point on the curve at ``time'' $s$. Note
  that the type of $\mathrm{quadB\acute{e}zier}$ tells us immediately
  that it preserves all changes of origin.
\end{example}

% Affine combination only provides us with an operation on points. We
% are also interested in operations on the vectors representing offsets
% between points. We now examine the correct types to assign to the
% vector space operations of addition of vectors, negation of vectors,
% multiplication by a scalar and the zero vector. The typings of these
% operations will make use of the abelian group structure of change of
% origin translations.

Vector operations are not invariant under change of origin. The
obvious type for vector addition is therefore:
\begin{displaymath}
  (+) : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
\end{displaymath}
(we write binary operators intended to be used infix in parentheses
when not appearing in infix position, following the Haskell
syntax.)

However, it is possible to give vector addition a more precise type
that shows how it is variant with change of origin:
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: \SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
\end{displaymath}
Intuitively, this type says that if the first input vector has been
displaced by $t_1$ and the second by $t_2$, then their sum is
displaced by $t_1 + t_2$. We can also negate vectors, yielding a
vector which points in the opposite direction. Negation negates
translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
\end{displaymath}
With the primitive operations of addition and negation of vectors, we
can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:\SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\tyPrim{vec}{t}$---we can compute their offset, which is a vector
expressed with respect to the null change of origin. The offset
operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
Note how the type of the result of $\mathrm{offset}$ simplifies from
$\tyPrim{vec}{t-t}$ to $\tyPrim{vec}{0}$. The algebraic structure on
the indexing theory induces type equalities.  The type of the vector
addition operation can be specialised to the case of moving a point by
a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5em}:\hspace{0.5em}}l}
    \mathrm{0} & \tyPrim{vec}{0}&
    (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
  \end{array}
\end{displaymath}

>>>>

We motivate our investigation of algebraically indexed types and their
relational interpretations by developing a novel type system for
programs that manipulate two-dimensional geometric data. Geometry is
rich with operations that are invariant under transformation: affine
operations are invariant under change of origin, vector space
operations are invariant under change of basis, and dot product is
invariant under orthogonal changes of basis. On the other hand, some
geometric operations are interestingly variant under transformation.
For example, cross products vary with scalings of the plane. We
can incorporate (in)variance information about geometric primitives
into type systems via algebraically indexed types.

\subsection{Origin Invariance and Representation Independence}\label{sec:oiri}

The $n$-tuple of numbers is the basic data structure for programs that
manipulate geometric data. Pairs, i.e., $2$-tuples, 
%The basic data structure used in programs that manipulate geometric data is
%the $n$-tuple of numbers. In the 2-dimensional case, tuples 
$\vec{v} = (x,y)$ serve double duty, representing
%are called upon to represent both 
both \emph{points}---offsets from some origin---and
\emph{vectors}---offsets in their own right.  Despite their common
representation, points and vectors are very different, and
distinguishing between them is the key feature of \emph{affine
  geometry} (see, for example, Chapter 2 of
\cite{gallier11geometric}). Nevertheless, computational geometry
libraries traditionally either leave it to the programmer to maintain
the distinction between points and vectors, or else use different
abstract types for points and vectors to enforce it.
%(one might say that the standard mathematical formulation of affine
%spaces uses the abstract types approach). 
In this paper we investigate a more sophisticated approach based on 
types indexed by change of origin transformations.

This approach regards the difference between points and vectors as a
change of data representation. For example, if $(0,0)$ and $(10,20)$
are two origins, then the tuple $(1,1)$ with respect to $(0,0)$ the
tuple $(11,21)$ with respect to $(10,20)$ represent \emph{the same
  point} because they have the same displacement from these two
origins, respectively.
%It is merely an artifact of the representation of points as pairs of
%numbers that they appear to be different.
This suggests that programs that manipulate %data representing points
should be invariant with respect to changes in the choice of
origin. Program sthat manipulate vectors, on the other hand, should
not be invariant under change of origin. Different vectors represent
different offsets, and the vector $(0,0)$ always represents the zero
offset.

Invariance under change of representation immediately recalls
Reynolds' fable about two professors teaching the theory of complex
numbers \cite{reynolds83types}. One professor represents complex
numbers using rectangular coordinates ($x + iy$), while the other
represents them using polar coordinates ($\alpha\cos\theta +
i\alpha\sin\theta$). Happily, after learning the basic operations on
complex numbers in the two representations, the two classes can
interact because the theory of complex numbers is invariant under the
choice of representation. Reynolds formalises the idea of invariance
under changes of representation as preservation of relations.  For
example, if a binary relation $R$ relates the rectangular and polar
representations of complex numbers, then a program that manipulates
complex numbers at a level of abstraction above their specific
representation should preserve $R$. %More precisely, if
%$f$ is a program that is invariant under the choice of representation
%of complex numbers,
%%taking complex numbers to complex numbers that respects invariance
%%under change of representation. Then 
%$c$ is a complex number in rectangular form, $c'$ is a complex number
%in polar form, and $R$ is a relation that relates the rectangluar and
%polar representations of complex numbers, then $(c,c') \in R$ implies
%$(f(c), f(c')) \in R$.

Reynolds' relational approach can be applied in the geometric setting
to show how quantifying over all changes of origin ensures the
invariance of programs under any particular choice of origin. For
this, we first define a family of binary relations on $\mathbb{R}^2$
that is indexed by changes of origin. Changes of origin are
represented by vectors in $\mathbb{R}^2$, and form a group
$\Transl{2}$ of translations under addition. The $\Transl{2}$-indexed
family of binary relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2
\times \mathbb{R}^2 \}_{\vec{t} \in \Transl{2}}$ is then defined by
%\begin{displaymath}
$R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}$.
%\end{displaymath}
We then consider a function $f$ that takes as input two tuples in
$\mathbb{R}^2$ and returns a single tuple in $\mathbb{R}^2$. We intend
that the tuples all represent points with respect to the same origin,
and that $f$ is invariant under the choice of origin.  Reynolds'
relational approach formalises this intention precisely: for any
$\vec{t} \in \Transl{2}$: % (i.e.~any change of origin), the function
%$f$
%should satisfy the following statement:
\begin{equation}\label{eq:f-preserve-rel-frame}
  \forall (\vec{v_1},\vec{v'_1}) \in R_{\vec{t}},
  (\vec{v_2},\vec{v'_2}) \in R_{\vec{t}}. (f(\vec{v_1}, \vec{v_2}),
  f(\vec{v'_1}, \vec{v'_2})) \in R_{\vec{t}}
\end{equation}
Unfolding the definition of $R_{\vec{t}}$ gives the equivalent formulation:
%\statementref{eq:f-preserve-rel-frame} is equivalent to: 
%for all $\vec{t} \in \Transl{2}$,
\begin{equation}\label{eq:f-invariant-frame}
  \forall \vec{v_1}, \vec{v_2}.\ f(\vec{v_1} + \vec{t},\vec{v_2} +
  \vec{t}) = f(\vec{v_1},\vec{v_2}) + \vec{t}.
\end{equation}
Thus, Reynolds' preservation of relations, when instantiated with the
family of relations $\{R_{\vec{t}}\}$, yields exactly the geometric
property of invariance under change of origin.

@@

Reynolds also showed how a type discipline can be used to establish
that (the denotational interpretations of) programs preserve
relations. For Reynolds, the type discipline of interest was that of
the polymorphic $\lambda$-calculus, which supports the construction of
new types by universal quantification over types.
%, in which new types can be
%constructed by universal quantification over all types.
In terms of relations, Reynolds interprets universal quantification
over types as quantification over binary relations between denotations
of types. By contrast, in our statements of geometric invariance in
\autoref{sec:type-system-geom-intro} we did not quantify over all
relations, but instead quantified over all changes of origin and used
a specific choice of origin to select a relation from the family
$\{R_{\vec{t}}\}$. This suggests introducing quantification over
changes of origin into the language of types. We use the notation
$\forall t \mathord: \SynTransl{2}. A$ for quantification over all
2-dimensional translations (i.e.,~choices of origin) $t$, and refer to
$\SynTransl{2}$ as the \emph{sort} of $t$. Note the difference in
fonts used to distinguish the semantic group $\Transl{2}$ from the
syntactic sort $\SynTransl{2}$. We use a similar convention
below, too.

Since the sort $\SynTransl{2}$ represents an abelian group, we can
combine its elements using the usual group operations. We write
operations additively, using $e_1 + e_2$ for the group operation, $-e$
for inverse and $0$ for the unit.  We also regard expressions built
from variables and the group operations up to the abelian group
axioms. For example, we regard $e_1 + (e_2 + e_3)$ and $(e_1 + e_2) +
e_3$ as equivalent.

Our language of types includes the unit type $\tyPrimNm{unit}$ and,
for all types $A$ and $B$, the function type $A \to B$, the sum type
$A + B$, and the tuple type $A \times B$. We also assume a primitive
type $\tyPrimNm{real}$, used to represent scalars. Although tuples of
real numbers represent points and vectors in geometric applications,
we cannot express this via the type
%central data structure in geometric applications is the tuple of real
%numbers for representing points and vectors, we cannot simply express
%this as the type 
$\tyPrimNm{real} \times \tyPrimNm{real}$. Indeed,
%because this type does not have the correct relational
%interpretation. (
two elements of type $\tyPrimNm{real}$ are related if and only if they
are equal and, by Reynolds' interpretation of tuple types, two
elements of $\tyPrimNm{real} \times \tyPrimNm{real}$ are also related
if and only if they are equal. But since this does not give the
correct relational interpretations for points and vectors, we
introduce a new type $\tyPrim{vec}{e}$, indexed by expressions $e$ of
sort $\SynTransl{2}$, to represent them.  The index $e$ represents the
displacement by change of origin of a point of this type.  Although we
have taken pains to distinguish geometric points and vectors, we use
the name $\tyPrimNm{vec}$ for both to recall the computer science
notion of vector as a homogeneous sequence of values with a known
length (in this case, 2).

As is standard for parametricity, every type has two interpretations:
an index-erasure interpretation that ignores the indexing expression,
and a relational interpretation as a binary relation on the
index-erasure interpretation. To give such interpretations for the
types $\tyPrim{vec}{e}$ and $\forall t\mathord:\SynTransl{2}.A$,
we assume for now that we can map each expression $e$ of sort
$\SynTransl{2}$ to an element $\sem{e}\rho$ of the group $\Transl{2}$
using some environment $\rho$ that interprets $e$'s free
variables. The (simplified) index-erasure and relation interpretations
of $\tyPrim{vec}{e}$ are:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e}} & \mathbb{R}^2
    \\ \rsem{\tyPrim{vec}{e}}{}\rho & R_{\sem{e}\rho} = \{
    (\vec{v},\vec{v'}) \sepbar \vec{v'} = \vec{v} + \sem{e}\rho \}
  \end{array}
\end{displaymath}
The index-erasure and relational interpetations of 
$\forall t\mathord:\SynTransl{2}.A$ are:
%is simply the interpretation of the
%underlying type $A$. The relational interpretation intersects the
%relational interpretations of the type $A$ under all extensions of the
%relational environment $\rho$:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\forall t\mathord:\SynTransl{2}.A} & \tySem{A}
    \\ \rsem{\forall t\mathord:\SynTransl{2}.A}{}\rho & \bigcap\{
    \rsem{A}{}{\rho'} \sepbar \rho' \in
    \extends{\rho}{t\mathord:\SynTransl{2}} \}
  \end{array}
\end{displaymath}
The exact definition of $\extends{\rho}{t\mathord:\SynTransl{2}}$
depends on which (possibly non-compositional) interpretation of
relational environments we use. For now we can think of environments
as mappings from free variables of sort $\SynTransl{2}$ to the group
$\Transl{2}$, and of extension as adding extra components to
mappings. The index-erasure and relational interpretations are given
formally in Sections~\ref{sec:index-erasure-semantics}
%the relational interpretation is in general more complex, due to the
%possibility of non-compositional interpretations of free index
%variables. The relational interpretation will be presented in
and~\ref{sec:relational-semantics}. %, respectively.

At the end of \autoref{sec:oiri} we considered functions $f :
\mathbb{R}^2 \times \mathbb{R}^2 \to \mathbb{R}^2$ that preserve all
changes of origin. This property of $f$ can be expressed in terms of
types by
%\begin{displaymath}
$  \mathrm{f} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t}
  \times \tyPrim{vec}{t} \to \tyPrim{vec}{t}$.
%\end{displaymath}
Spelling out the relational interpretation of this type using the
definitions above and the standard relational interpretations for
tuple and function types, we recover
\statementref{eq:f-invariant-frame} exactly.

@@

Invariance under change of origin is the key feature of affine
geometry, whose central operation is the affine combination of points:
$\lambda_1\vec{v_1} + \lambda_2\vec{v_2}$, where $\lambda_1 +
\lambda_2 = 1$.  Geometrically, this can be interpreted as
interpolation between the points represented by $\vec{v_1}$ and
$\vec{v_2}$.  We add affine combination of points to our calculus as
follows: % typing and intended denotation:for
\begin{displaymath}
  \begin{array}{l}
    \mathrm{affComb} :\forall t \mathord:
    \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrimNm{real} \to
    \tyPrim{vec}{t} \to \tyPrim{vec}{t} \\ 
    \sem{\mathrm{affComb}}\ \vec{v_1}\ r\ \vec{v_2} = (1-r)\vec{v_1} +
    r\vec{v_2} 
\end{array}
\end{displaymath}
It can be verified by hand that %the intended denotation
$\sem{\mathrm{affComb}}$ has the invariance property specified by its
type. {\bf Vague. Say which invariance property.}
% By defining the primitive function $\mathrm{affComb}$ to take a
% single $\tyPrimNm{real}$ parameter $t$, we can easily ensure that we
% are taking the affine combination of two representatives of points,
% and not just an arbitrary linear combination.

\begin{example}
  The evaluation of quadratic B\'{e}zier curves (B\'{e}zier curves
  with two endpoints and a single control point) can be expressed
  using the affine combination primitive as follows: % and three steps
  % of De Casteljau's algorithm:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{quadB\acute{e}zier} : \forall t
      \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to
      \tyPrim{vec}{t} \to \tyPrimNm{real} \to \tyPrim{vec}{t}
      \\ \mathrm{quadB\acute{e}zier}\ [t]\ p_0\ p_1\ p_2\ s = \\ \quad
      \mathrm{affComb}\ [t]\ (\mathrm{affComb}\ [t]\ p_0\ s\ p_1)\ s\ (\mathrm{affComb}\ [t]\ p_1\ s\ p_2)
      \\
    \end{array}
  \end{displaymath}
  For two endpoints $p_0$ and $p_2$, a control point $p_1$, and $s \in
  [0,1]$, an application
  $\mathrm{quadB\acute{e}zier}\ p_0\ p_1\ p_2\ s$ gives the point on
  the curve at ``time'' $s$.  The type of
  $\mathrm{quadB\acute{e}zier}$ immediately tells us that it preserves
  all changes of origin.
\end{example}
% Affine combination only provides us with an operation on points. We
% are also interested in operations on the vectors representing offsets
% between points. We now examine the correct types to assign to the
% vector space operations of addition of vectors, negation of vectors,
% multiplication by a scalar and the zero vector. The typings of these
% operations will make use of the abelian group structure of change of
% origin translations.

The obvious type for vector addition is
%\begin{displaymath}
$  (+) : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}$.
%\end{displaymath}
%(we write binary operators intended to be used infix in parentheses
%when not appearing in infix position, following the Haskell
%syntax.)
But we can reflect the fact that $(+)$ is not invariant under change
of origin by giving it a more precise type that reflects how it varies
with change of origin:
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: \SynTransl{2}.\ \tyPrim{vec}{t_1}
  \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
\end{displaymath}
Intuitively, this type says that if the first input vector has been
displaced by $t_1$ and the second by $t_2$, then their sum is
displaced by $t_1 + t_2$. We can also negate vectors, yielding a
vector which points in the opposite direction. Negation negates
translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
\end{displaymath}
Finally, with the primitive operations of addition and negation of
vectors we can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:\SynTransl{2}.\ \tyPrim{vec}{t_1}
    \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2}
    \\ (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

Given two points that are invariant with respect to the same change of
origin---i.e.~two values of type $\tyPrim{vec}{t}$---we can use
subtraction to compute their offset:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t
    \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to
    \tyPrim{vec}{0} \\ \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
The result is a vector expressed with respect to the null change of
origin: note how the algebraic structure on the indexing theory
induces type equalities that can be used to simplify the type of the
result of $\mathrm{offset}$ from $\tyPrim{vec}{t-t}$ to
$\tyPrim{vec}{0}$. The type of $(+)$ can also be specialised to the
case of moving a point by a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t
    \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to
    \tyPrim{vec}{t} \\ \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, namely
$\mathrm{0} : \tyPrim{vec}{0}$ for
the zero vector and $(*) : \tyPrimNm{real} \to
    \tyPrim{vec}{0} \to \tyPrim{vec}{0}$ for
multiplication by a scalar, do not describe any
interesting effects on translations.
%\begin{displaymath}
%  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5%em}:\hspace{0.5em}}l}
%    \mathrm{0} & \tyPrim{vec}{0}& (*) & \tyPrimNm{real} \to
%    \tyPrim{vec}{0} \to \tyPrim{vec}{0}
%  \end{array}
%\end{displaymath}
